---
title: Mixture models
author: 
  name: Matthew Stephens
  affiliation: University of Chicago
date: January 12, 2026
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: readable.tex
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
---

See [here][pdf_version] for a PDF version of this vignette.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,fig.align = "center")
```

Prerequisites 
=============

You should be familiar with basic probability, including the law of
total probability, the notion of distibution and density, and with
standard distributions (particularly the Gamma distribution).

Overview
========

This vignette introduces the idea of a mixture model. These models are
widely used in statistics to model data where observations come from a
"mixture" of two or more different distributions. This vignette
introduces the basic idea of a mixture, its density/mass function, and
terminology such as *mixture proportions*, *mixture components* and
*latent variable representation.*

Example
=======

We begin this vignette with an example.  A medical screening test for
a disease involves measuring the concentration ($X$) of a protein in
the blood. In normal individuals, $X$ has a Gamma distribution with
mean 1 and shape 2 (so the scale parameter is 0.5, as scale =
mean/shape).  In diseased individuals, the protein becomes elevated,
and $X$ has a Gamma distribution with mean 2 and shape 2 (so scale =
1).

Suppose that in a particular population, 70% of individuals are normal
and 30% are diseased. What will be the overall distribution of the
protein levels in this population? How could you simulate from this
distribution?

Answer
------

### The overall distribution

Let $X$ denote the protein concentration of a randomly chosen
individual from the population. Let $Z$ denote whether the
randomly-chosen individual is normal ($Z=1$) or diseased ($Z=2$). Here
we assume that $Z$ is not observed; it has been introduced to help
with notation and derivations.

By the law of total probability, we can write the density of $X$ as
$$
p(x) = \Pr(Z = 1) \, p(x \mid Z = 1) +
\Pr(Z=2) \, p(x \mid Z = 2).
$$
In words, this represents
$$
p(x) = \Pr(\text{normal}) \, p(x \mid \text{normal}) +
\Pr(\text{diseased}) \, p(x \mid \text{diseased}).
$$

From the information given, we know that $\Pr(\text{normal})=0.7$ and
$\Pr(\text{diseased}) = 0.3$. We also know $p(x \mid \text{normal})$
and $p(x \mid \text{diseased})$ are each given by the density of a
Gamma distribution. So we can write
$$
\textstyle
p(x) = 0.7 \, \mathrm{Gamma}(x; 0.5, 2) + 0.3 \, \mathrm{Gamma}(x; 1, 2),
$$
where $\mathrm{Gamma}(x; a, b)$ denotes the density of a Gamma
distribution with scale $a$ and shape $b$.

This distribution is an example of a "mixture distribution" (in
particular, it is a "mixture of two Gamma distributions").  Here we
plot the density of this mixture distribution (solid blue), as well as
the densities of the individual distributions that were combined to
make the mixture (the dashed black line for "normal", the dashed red
line for "diseased"). In mixture terminology, these individual
distributions are called the "component distributions".

```{r gammamix, fig.height=3.5, fig.width=4}
x <- seq(0,10,length.out = 100)
plot(x,
     0.7 * dgamma(x,scale = 0.5,shape = 2) +
     0.3 * dgamma(x,scale = 1,shape = 2),
     col = "blue",type = "l",lwd = 1.5,
     xlab = "protein concentration",
     ylab = "density",ylim = c(0,0.8))
lines(x,dgamma(x,scale = 0.5,shape = 2),type = "l",col = "black",
      lwd = 1,lty = "dashed")
lines(x,dgamma(x,scale = 1,shape = 2),type = "l",col = "red",lwd = 1,
      lty = "dashed")
```

### Simulation

One nice thing about mixture models is that they are easy to
simulate from. The trick is to simulate both $X, Z$ from the joint
distribution $p(X, Z)$, then simply ignore $Z$. This ensures
that $X$ comes from its marginal distribution $p(X)$.

Simulating from $p(X, Z)$ can be achieved by a two-stage process: (1)
simulate $Z \sim p(Z)$, (2) simulate $X \mid Z \sim p(X \mid Z)$, both
of which are easy.  The following code illustrates this idea by
simulating 10,000 samples from the mixture, then plotting a
histogram of the samples. As you can see, the histogram closely matches
the mixture density.

```{r gammamix-sim, fig.height=3.5, fig.width=4.5}
n <- 10000
x <- rep(0,n)
shape <- c(2,2)
scale <- c(0.5,1)
for (i in 1:n) {
  if (runif(1) < 0.7) 
    z <- 1
  else
    z <- 2
  x[i] <- rgamma(1,scale = scale[z],shape = shape[z])
}
hist(x,breaks = 64,xlim = c(0,10),probability = TRUE,main = "")
xvec <- seq(0,10,length.out = 100)
lines(xvec,
      0.7 * dgamma(xvec,scale = 0.5,shape = 2) +
      0.3 * dgamma(xvec,scale = 1,shape = 2),
      lwd = 1.5,col = "blue")
```

Note that we set `probability = TRUE` when creating the histogram in
the above code so that it is scaled like a density (i.e., area =
1). This makes it easy to compare with the theoretical density.

Exercises
=========

The following exercises are designed to help you generalize the ideas
in example above to other settings.

## 1. Other mixture component distributions

In the above example, there was people sampled from a population that
is 50% male and 50% female, where the males have heights that are
normally distributed with mean 70 in and standard deviation 3 in, and
the females have heights that are normally distributed with mean 64.5
in and standard deviation 2.5 in. Write the density of the mixture
model. Identify the mixture proportions and the mixture component
densities. Plot the mixture density and the component densities in a
plot similar to the one in the example above.

## 2. More than two distributions

Similarly, there is nothing that limits us to mixing just two
distributions — you can mix any number of distributions
together. This exercise illustrates this idea.

Suppose that in the protein concentration example above, the
population consists of males and females, who have different rates of
disease, and also different protein distributions. So now there are
four groups: "male, diseased"; "male, normal"; "female, diseased"; and
"female, normal".  Making whatever assumptions you want to make about the
component distributions (say what they are), and about the relative
frequency of each group in the population (again, say what they are),
write out a mixture distribution that could represent this
situation. Plot the components of your assumed mixture and the mixture
density.

## 3. Discrete data

The above example involves a mixture of two continuous distributions
(Gamma distributions) and so the mixture distribution is also
continuous. However, mixtures of discrete distributions work in the
same way: you just use probability mass functions instead of
probability density functions.

For example, let $X$ denote the number of molecules of a particular
gene in a cell randomly drawn from some population of cells. Assume
that the population contains three cell types, in proportions
0.2, 0.4, 0.4. Further suppose that in each cell type the number of
molecules follows a Poisson distribution with mean parameter $\lambda
= 2, 5, 10$, respectively. The distribution of $X$ is therefore a
mixture of three Poisson distributions. Can you write down its
probability mass function?

The general case
================

Putting the above ideas together, we can write the density for a
mixture of $K$ continuous distributions that have densities $f_1,
\dots, f_K$ in proportions $\pi_1, \dots, \pi_K$. Such a mixture would
have density
$$
p(x) = \sum_{k=1}^K \pi_k f_k(x).
$$
Exactly the same equation holds for a mixture of $K$ discrete
distibutions, but with $p(x)$ and $f_k$ representing probability mass
functions instead of densities.

Some important terminology
--------------------------

+ This is called a *mixture distribution* (or mixture model, or just
  mixture) with $K$ *components*. (Sometimes it is called a *finite
  mixture* because one can also further generalize the ideas to an
  uncountably infinite number of components!)

+ The $f_1, \dots, f_K$ are called the *component densities* (or
  *component distributions*). So $f_1$ is the density of component 1,
  and $f_k$ is the density of component $k$.

+ The $\pi_1, \dots, \pi_K$ are called the *mixture proportions*. Of
  course, we must have $\pi_k \geq 0$ and $\sum_{k=1}^K \pi_k =1$.

+ The unobserved random variable $Z$ is sometimes referred to as the
  "component of origin" or the "component that gave rise to" the
  observation $X$. If we have $n$ observations $X_1,\dots,X_n$ from a
  mixture model it is common to use $Z_i$ to denote the component that
  gave rise to $X_i$.

+ Introducing unobserved variables to help with computations
  or derivations is a common trick that is used beyond mixture models.
  This trick is sometimes called *data augmentation*.  The unobserved
  random variables are sometimes called "hidden variables" or "latent
  variables".  Representing the mixture model
  $$
  p(x) = \sum_{k=1}^K
  \pi_k f_k(x)
  $$
  by the two-stage process
  $$
  \begin{aligned}
  p(Z=k) &= \pi_k \\
  p(x \mid Z = k) &= f_k(x), \quad k = 1, \ldots, K
  \end{aligned}
  $$ is called the *latent variable representation* of
  the mixture model.

[pdf_version]: https://github.com/pcarbo/fiveMinuteStats/blob/master/docs/mixture_models_01.pdf
