---
title: Introduction to bivariate normal distribution
author: 
  name: Matthew Stephens
  affiliation: University of Chicago
date: February 10, 2026
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: readable.tex
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
---

See [here][pdf_version] for a PDF version of this vignette.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,fig.align = "center")
```

# Prerequisites

You need to have basic familiarity with univariate normal
distribution, and understand the [basic property][norm_linear_comb]
that linear combinations of normals are also normal.

# Motivating example

Suppose that $Z_1, Z_2$ are independent standard normal $N(0,1)$, and
define $X_1 = Z_1 + 0.1 Z_2$, $X_2 = Z_1 - 0.1 Z_2$. What is the
joint distribution of $X_1, X_2$?

We know from the [basic property][norm_linear_comb] that $X_1$ will be
univariate normal, and that $X_2$ will be univariate normal. However,
they will not necessarily be independent because $Z_1$ and $Z_2$ were
used to compute both. Indeed, you can see that $X_1$ and $X_2$ might
both be expected to be close to $Z_1$ (because the 0.1 multiplier on
$Z_2$ is "relatively small"). So when $X_1$ is $X_2$ should also be
big, and when $X_1$ is small $X_2$ should also be small.

The following code illustrates this: the histograms illustrate both
$X_1$ and $X_2$ are normal, and the scatterplot of $X_1$ and $X_2$
shows they are correlated (and the sample correlation is
approximately 0.98).

```{r bivariate-example, fig.height=2.5, fig.width=6.5}
Z1 <- rnorm(1000)
Z2 <- rnorm(1000)
X1 <- Z1 + 0.1*Z2
X2 <- Z1 - 0.1*Z2
par(mfcol = c(1,3))
hist(X1,breaks = 32)
hist(X2,breaks = 32)
plot(X1,X2,main = "scatterplot of (X1,X2)",ylim = c(-4,4),asp = 1)
cor(X1,X2)
```

The bivariate normal distribution
=================================

In fact the answer to the question "what is the joint distribution of
$X_1,X_2$" is they have a "bivariate normal distribution". Thus the
scatterplot shown above shows a scatterplot of 1,000 samples from a
bivariate normal distribution.  The prefix "bi" means two, referring
to the fact that here we are looking at two variables, $X_1$ and
$X_2$. The ideas here can be extended to more variables, and the
resulting distribution is called the "multivariate normal" (multi =
two or more). The bivariate normal is a special case of the
multivariate normal.

Mean and covariance matrix
--------------------------

The bivariate normal distribution has 5 parameters: two means (for
$X_1$ and $X_2$), two variances (for $X_1$ and $X_2$) and the
covariance between $X_1$ and $X_2$. It is usual to write the mean
parameter as a vector $\mu$ and the variance and covariance parameters
as a $2 \times 2$ symmetric matrix $\Sigma$ where the diagonal
elements of $\Sigma$ contain the variances and the off-diagonal
elements contain the covariance.  $\Sigma$ is called the "covariance
matrix" or sometimes the "variance-covariance matrix".

General construction
--------------------

Suppose $Z_1, Z_2$ are independent random variables each with a
standard normal distribution $N(0,1)$. Let $Z$ denote the vector
$(Z_1, Z_2)$, let $A$ be any $2 \times 2$ matrix, and $\mu$ be any
$r$-vector. Then the vector $X = AZ + \mu$ has a bivariate normal
distribution with mean $\mu$ and variance-covariance matrix
$\Sigma = AA^T$. (Here, $A^T$ means the transpose of the matrix $A$.) We
write $X \sim N_2(\mu, \Sigma)$.

Example 
=======

We can redo the example above with vector and matrix notation, $\mu =
(0,0)^T$ and $A = \left[\begin{array}{cc} 1 & 0.1 \\ 1 & -0.1
\end{array}\right]$. Here for clarity we just simulate a single sample
instead of 1,000:

```{r bivariate-example-matrix-vector}
mu <- c(0,0)
A <- rbind(c(1,0.1),c(1,-0.1))
A
z <- rnorm(2)
x <- mu + A %*% z
x
```

It should be clear from the above that in our example the mean is
$\mu = (0,0)$.  What is the covariance matrix $\Sigma$? We can compute
it from the formula $\Sigma = AA^T$:

```{r cov-matrix}
Sigma <- A %*% t(A)
Sigma
```

[pdf_version]: https://github.com/pcarbo/fiveMinuteStats/blob/master/docs/mvnorm_00.pdf
[norm_linear_comb]: https://pcarbo.github.io/fiveMinuteStats/norm_linear_comb.html
