---
title: Multivariate normal
author: 
  name: Matthew Stephens
  affiliation: University of Chicago
date: February 10, 2026
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: readable.tex
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
---

See [here][pdf_version] for a PDF version of this vignette.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,fig.align = "center")
```

# Prerequisites

You need to know what a univariate normal distribution is, and basic
properties such as the fact that linear combinations of normals are
also normal. You also need to know the basics of matrix algebra
(e.g., matrix multiplication, matrix transpose).

Definition
==========

There are several equivalent ways to define a multivariate normal, but
perhaps the most succinct and elegant is this one, which I took from
Wikipedia: "a random vector is said to be $r$-variate normally
distributed if every linear combination of its $r$ components has a
univariate normal distribution". The special case $r = 2$ is usually
referred to as the "bivariate normal" distribution. And the
terminology "$r$-dimensional multivariate normal" means the same as
$r$-variate normal.

Perhaps more insight can be obtained by specifying one way to simulate
from a multivariate normal distribution. Suppose $Z_1, \dots, Z_n$ are
independent random variables each with a standard normal distribution
$N(0,1)$. Let $Z$ denote the vector $(Z_1, \dots, Z_n)$, and let $A$
be any $r \times n$ matrix, and $\mu$ be any $r$-vector. Then the
vector $X = AZ + \mu$ has an $r$-dimensional multivariate normal
distribution with mean $\mu$ and variance-covariance matrix $\Sigma :=
AA^T$. (Here, $A^T$ means the transpose of the matrix $A$.) We write
$X \sim N_r(\mu, \Sigma)$.

Note that $E(X_j) = \mu_j$ and $\mathrm{Cov}(X_i,X_j) = \Sigma_{ij}$.

When we simulate $X$ this way, any linear combination of the elements
of $X$ is a linear combination of $Z$, and so is univariate normal
because any linear combination of univariate normals is univariate
normal. That is, when we simulate $X$ this way it satisfies the
definition above for being $r$-variate normal.

Example
=======

Suppose we define $X_1 = Z_1 + Z_2$ and $X_2 = Z_1 + Z_3$ where
$Z_1,Z_2$ and $Z_3$ are independent standard normal variables. Then $X
= AZ$ where $A = \begin{pmatrix} 1 & 1 & 0 \\ 1 & 0 & 1
\end{pmatrix}$, and so $X \sim N_2 (0,\Sigma)$ where $\Sigma = AA^T =
\begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}$.

Here is some code to simulate from this distribution:

```{r sim}
Z <- rnorm(3)
A <- rbind(c(1,1,0),c(1,0,1))
Sigma <- A %*% t(A)
Sigma
X <- A %*% Z
```

And if we want to simulate lots of examples, we can do this lots of
times. The following code simulates from this distribution 1m000 times
and plots the points.

```{r more-sims, fig.height=3.5, fig.width=3}
X <- matrix(0,2,1000)
for(i in 1:1000) {
  Z <- rnorm(3)
  X[,i] <- A %*% Z
}
plot(X[1,],X[2,],main = "bivariate normal",asp = 1,
     xlab = "x1",ylab = "x2",xlim = c(-5,5),ylim = c(-5,5))
```

And we can check that the sample covariances are close to the
theoretical values:

```{r check-cov}
cov(t(X))
```

General algorithm 
=================

From the above, we can see that you can simulate from a multivariate
normal with any mean $\mu$ and variance covariance matrix $\Sigma$
provided that $\Sigma$ can be written as $\Sigma = AA^T$ for some
matrix $A$. This turns out to be possible if and only if $\Sigma$ is
what is known as a "positive semi-definite" matrix—that is, a
symmetric matrix with non-negative eigenvalues.

Given a positive semi-definite matrix $\Sigma$, there are multiple ways
to find a matrix $A$ such that $\Sigma = AA^T$ (and indeed multiple
matrices $A$ that obey this). However, for our purposes here we only
need one way, and here we use the "Cholesky decomposition", which
finds a unique lower triangular matrix $L$ such that $LL^T =
\Sigma$. Here's an illustration of the Cholesky decomposition (the R
function `chol()` finds an upper triangular matrix, so we transpose it
here to make it lower triangular).

```{r chol}
Sigma <- rbind(c(2,1),c(1,2))
L <- t(chol(Sigma))
L
L %*% t(L)
```

We can use this to generate *any* multivariate normal. Here we
use it to generate a bivariate normal with covariance matrix 
$$
\begin{pmatrix}
1.0 & 0.9 \\
0.9 & 1.0
\end{pmatrix}.
$$

```{r example-with-chol, fig.height=3.5, fig.width=3}
my_rmvnorm <- function (mu, Sigma) {
  r <- length(mu)
  L <- t(chol(Sigma)) 
  Z <- rnorm(r)
  return(L %*% Z + mu)
}
X <- matrix(0,2,1000)
S <- rbind(c(1.0,0.9),c(0.9,1.0))
for (i in 1:1000)
  X[,i] <- my_rmvnorm(c(0,0),S)
plot(X[1,],X[2,],main = " bivariate normal",asp = 1,xlab = "x1",ylab = "x2")
```

**Note:** If you want to actually generate multivariate normal random
variables in R you should use a more carefully written function like
"rmvnorm" from the mvtnorm package.

[pdf_version]: https://github.com/pcarbo/fiveMinuteStats/blob/master/docs/mvnorm.pdf
