---
title: Linear combinations of independent normals are normal
author: 
  name: Matthew Stephens
  affiliation: University of Chicago
date: February 9, 2026
output:
  pdf_document:
    keep_tex: false
    latex_engine: pdflatex
    template: readable.tex
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
spacing: single
graphics: yes
endnote: no
---

See [here][pdf_version] for a PDF version of this vignette.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,fig.align = "center")
```

Prerequisites
=============

Basic familiarity with the univariate normal distribution.

A statement of the basic property
=================================

The simple goal of this vignette is to introduce a basic property of
the (univariate) normal distribution: that linear combinations of
independent normal variables are also normal.

Formally, suppose $Z_1$ and $Z_2$ represent independent, normally
distributed random variables. Then for any scalars $a$ and $b$, the
linear combination
$$
X := aZ_1 + bZ_2
$$ 
is also (univariate) normal.

Also, by basic properties of expectation and variance, $E(X) =
aE(Z_1) + bE(Z_2)$ and $\mathrm{Var}(X) = a^2 \mathrm{Var}(Z_1) + b^2
\mathrm{Var}(Z_2)$.

Example
=======

The following code provides a visual illustration of this idea with
$a = 2$ and $b = 3$, but it holds for any $a$ and $b$.

First we sample some values of $X$ by randomly generating $Z_1$ and
$Z_2$, and computing $X = aZ_1 + bZ_2$:

```{r sim-x}
Z1 <- rnorm(1000)
Z2 <- rnorm(1000)
a <- 2
b <- 3
X <- a*Z1 + b*Z2 
```

The property says that the samples of $X$ look normal. A quick
histogram and qqplot suggest it does. (Of course, this is not a
proof that the property holds; it is just an illustration of the
idea.)

```{r plots, fig.height=3.5, fig.width=3.5}
hist(X)
qqnorm(X)
```

Addendum: Stable distributions
==============================

If you are curious by nature, you might now ask: is the normal
distribution the only distribution that satisfies this property?  The
answer is "no". For example, $t$ distributions also satisfy this
property. Distributions that satisfy this property are called "stable"
distributions. You can read more [here](wikipedia-stable-dist).

[pdf_version]: https://github.com/pcarbo/fiveMinuteStats/blob/master/docs/norm_linear_comb.pdf
[wikipedia-stable-dist]: https://en.wikipedia.org/wiki/Stable_distribution
